---
title: "Portfolio Project"
author: "Jay Gohner"
date: "October 10, 2019"
output: html_document
---


INTRODUCTION:
  I want to see if there are any relations between the levels of education that a certain state in the US has
  compared to it's rates of violent crimes. I believe that the more educated a population, the more access that
  population has to higher education, the less likely there is to be incidents of violent crime. Maybe information
  like this could help to show with hard numbers why we need to take education more seriously in this country.
  
  The initial dataset I will start with is from the USDA and it contains census bureau figures from 1970, 1980, 1990
  and 2000 as well as the 2013 - 2017 American Community Survey 5 year average.
  
  The data was quite vast, including data points from each county in America, so I made the conscious decision to
  focus only on states. I also decided to drop the remaining columns for 2013-2017 as I felt they didnt fit with the
  decade format the other information was in.
  
  Finding this source was actually more difficult than I had initially assumed, apparently the current
  administration been quietly removing many datasets. Thankfully the USDA has not been big on complying with these
  "requests" and I was able to obtain this dataset that I feel gives me a good start in the right direction for my
  project. The next steps will involve finding reliable sources of violent crime rates by county in these same
  years. I think another dataset of money spent on higher education and perhaps concentration of schools in areas
  could also provide me with more insights.
  
  The original column names are as follows:
      State - this contains the state the observation takes place in
      Area name - this contains the county the observation takes place in
      Less than a high school diploma, 1970 - number of adults who have less than a high school diploma
      High school diploma only, 1970 - number of adults who have who have a high school diploma
      Some college (1-3 years), 1970 - number of adults who have who have some college education
      Four years of college or higher, 1970 - number of adults who have completed 4 or more years of college
      Percent of adults with less than a high school diploma, 1970 - percentage of adults
      Percent of adults with a high school diploma only, 1970 - percentage of adults
      Percent of adults completing some college (1-3 years), 1970 - percentage of adults
      Percent of adults completing four years of college or higher, 1970 - percentage of adults
      These columns are then repeated for 1980, 1990 and 2000
      Less than a high school diploma, 2013-17 - 
      High school diploma only, 2013-17 - 
      Some college or associate's degree, 2013-17 
      Bachelor's degree or higher, 2013-17 
      Percent of adults with less than a high school diploma, 2013-17
      Percent of adults with a high school diploma only, 2013-17
      Percent of adults completing some college or associate's degree, 2013-17
      Percent of adults with a bachelor's degree or higher, 2013-17
  
  A note on the data: for 1970 and 1980, the share of adults with less than high school includes those who had not
  completed the 12th grade. In 1990, 2000, and 2013-17 the share includes those who did not receive a high school
  diploma or its equivalent (such as a GED).
  
  I will be renaming the column names in the original dataset and dropping the first column which is some sort of
  unique key id along with four other unique key ids that are not relevant.
  
  Certain functions and special packages in R can help us organize this data. They are loaded in as libraries. I use the
  suppressMessages function to make the print out a bit cleaner.
```{r}
suppressMessages(library("tidyverse"))
suppressMessages(library("tidyr"))
suppressMessages(library("dplyr"))
suppressMessages(library("readr"))
suppressMessages(library("stringr"))
suppressMessages(library("readxl"))
```

After loading all necessary libraries, we will import the Education.csv dataset using read_csv(). This function reads in a
csv and makes it into a tibble. A tibble can be thought of as something similar to an excel spreadsheet.
```{r}
eduLevelsbyCounty <- read_csv("Education.csv")
```

This data gives the numbers and percentages for adults 18 and older and the level of education they obtained by county in
the USA. The column names were long and wordy so I cleaned them up by accessing thw column names and then assigning them
new ones. I then use the select() function to drop some unique identifier columns that serve no purpose to this project.

```{r}
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "State"] <- "state"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Area name"] <- "area_name"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Less than a high school diploma, 1970"] <- "Less Than Diploma, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "High school diploma only, 1970"] <- "Diploma, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Some college (1-3 years), 1970"] <- "AA or more, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Four years of college or higher, 1970"] <- "BA or more, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with less than a high school diploma, 1970"] <- "%Less Than Diploma, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a high school diploma only, 1970"] <- "% Diploma, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults completing some college (1-3 years), 1970"] <- "% AA or more, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults completing four years of college or higher, 1970"] <- "% BA or more, 1970"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Less than a high school diploma, 1980"] <- "Less Than Diploma, 1980"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "High school diploma only, 1980"] <- "Diploma, 1980" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Some college (1-3 years), 1980"] <- "AA or more, 1980" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Four years of college or higher, 1980"] <- "BA or more, 1980" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with less than a high school diploma, 1980"] <- "% Less Than Diploma, 1980" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a high school diploma only, 1980"] <- "% Diploma, 1980" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults completing some college (1-3 years), 1980"] <- "% AA or more, 1980" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults completing four years of college or higher, 1980"] <- "% BA or more, 1980"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Less than a high school diploma, 1990"] <- "Less Than Diploma, 1990"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "High school diploma only, 1990"] <- "Diploma, 1990" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Some college or associate's degree, 1990"] <- "AA or more, 1990" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Bachelor's degree or higher, 1990"] <- "BA or more, 1990" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with less than a high school diploma, 1990"] <- "% Less Than Diploma, 1990" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a high school diploma only, 1990"] <- "% Diploma, 1990" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults completing some college or associate's degree, 1990"] <- "% AA or more, 1990" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a bachelor's degree or higher, 1990"] <- "% BA or more, 1990"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Less than a high school diploma, 2000"] <- "Less Than Diploma, 2000"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "High school diploma only, 2000"] <- "Diploma, 2000" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Some college or associate's degree, 2000"] <- "AA or more, 2000" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Bachelor's degree or higher, 2000"] <- "BA or more, 2000" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with less than a high school diploma, 2000"] <- "% Less Than Diploma, 2000" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a high school diploma only, 2000"] <- "% Diploma, 2000" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults completing some college or associate's degree, 2000"] <- "% AA or more, 2000" 
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a bachelor's degree or higher, 2000"] <- "% BA or more, 2000"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Less than a high school diploma, 2013-17"] <- "lessHSD13_17"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "High school diploma only, 2013-17"] <- "HSD13_17"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Some college or associate's degree, 2013-17"] <- "someCollegeorAA13_17"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Bachelor's degree or higher, 2013-17"] <- "bachelorsOrMore13_17"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with less than a high school diploma, 2013-17"] <- "percLessHSD13_17"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a high school diploma only, 2013-17"] <- "percHSD13_17"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults completing some college or associate's degree, 2013-17"] <- "percAA13_17"
colnames(eduLevelsbyCounty)[colnames(eduLevelsbyCounty) == "Percent of adults with a bachelor's degree or higher, 2013-17"] <- "percBA13_17"

eduLevelsbyCounty <- select(eduLevelsbyCounty, 2, 3, 8:39)

whole_nums_edu_levels <- select(eduLevelsbyCounty, 1:6, 11:14, 19:22, 27:30)
percents_edu_levels <- select(eduLevelsbyCounty, 1:2, 7:10, 15:18, 23:26, 31:34)
```

I decided to graph the US as a whole and show the trends of adult education levels from 1970 to 2000. This was
accomplished by using the filter function and piping operator to make two new dataframes. One contains the whole
number amount of adults who have obtained one of the four education levels and the second is the percentage value
of that same statistic.
I then make the data into more readable tibble using pivot_longer and then to provide a graph by year I use
separate to brek up the education level and the year.
```{r}
whole_nums_USA <- whole_nums_edu_levels %>%
  filter(state == "US")

percents_USA <- percents_edu_levels %>%
  filter(state == "US")

nums_df <- whole_nums_USA %>% pivot_longer(cols=c(3:ncol(whole_nums_USA)) )

nums_df <- nums_df %>% separate(name, into=c("name", "year"), sep=", ")

percents_df <- percents_USA %>% pivot_longer(cols=c(3:ncol(percents_USA)) )

percents_df <- percents_df %>% separate(name, into=c("name", "year"), sep=", ")
```

Now using my new dataframes with the information I want to visualize, I use ggplot to make a plot.
```{r}

library(ggplot2)
#make line graph for whole nums
whole<-ggplot(nums_df, aes(x=year, y=value, group=name, color=name)) + scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
  geom_line()

print(whole)

percents<-ggplot(percents_df, aes(x=year, y=value, group=name, color=name)) + scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
  geom_line()

print(percents)
```
We see that the overall trend is an increase in adults obtaining higher levels of education. What we hope to see 
once the new datasets regarding rates of violent crime are introduced is an overall decrease in incidents of violent
crime.
I also hope to see that the rates of violent crime are decreasing with the increasing of adults obtaining higher
levels of education.